---
Title: Analysis of Wildlife Strikes in Commercial Aviation
Subtitle: Practicum I CS5200
Author: Kuan Chun Chiu
Author_email: chiu.kua@northeastern.edu
Date: 2024 Fall Semester
---

# Analysis of Wildlife Strikes in Commercial Aviation
## Practicum I CS5200
### Author: Kuan Chun Chiu
### Date: 2024 Fall Semester

## Question 3. Connect to MySQL Database
```{r connect_to_db, echo=T, warning=F}
library(DBI)
library(RMySQL)

# Define connection parameters
host = "derekchiu-mysql-avien-service-aircraft-birds-strike-analysis.i.aivencloud.com"
user = "avnadmin"
password = "AVNS_l7gYA424n1NtdnepYjq"

# Connect to MySQL
conn = dbConnect(MySQL(), user=user, password=password, host=host,                   port=26901)

# Create a database in MySQL
dbExecute(conn, "CREATE DATABASE IF NOT EXISTS aircraft_db")

# Connect to the aircraft_db database
dbDisconnect(conn)
conn = dbConnect(MySQL(), user=user, password=password, host=host,
                 port=26901, dbname="aircraft_db")
conn
```
### Temporarily Disabled the Foreign Key Constraint When Creating Tables
```{r disable_fk_constraint, echo=T, warning=F}
# Disable the foreign key checking constraint, so parents table can be created before child tables
disable_fk = "
SET FOREIGN_KEY_CHECKS = 0;
"
dbExecute(conn, disable_fk)
```
### Drop All Tables, Stored Procedures, and Triggers Within the Database
```{r, clean_up_database, echo=T, warning=F}
# Drop the tables first so later when the code ran again during kitting, the tables can be re-created and data can be re-inserted into them
for (table in dbListTables(conn)){
  if (table == "wildlife.strikes"){
    table = "`wildlife.strikes`"
  }
  dbExecute(conn, paste0("DROP TABLE IF EXISTS ", table))
}

# Drop the stored procedure
dbExecute(conn, "DROP PROCEDURE IF EXISTS update_airport_code")

# Drop the trigger
dbExecute(conn, "DROP TRIGGER IF EXISTS check_airportcode_update")
```
## Question 4. Create Tables in the Database
```{r create_tables, echo=T, warning=F}
# Part A, C
create_flights = "
CREATE TABLE IF NOT EXISTS flights (
  fid INTEGER PRIMARY KEY,
  date DATE,
  originAirport INTEGER,
  airlineName TEXT,
  aircraftType TEXT,
  isHeavy TINYINT(1) DEFAULT 0,
  FOREIGN KEY (originAirport) REFERENCES airports (aid)
);
"
# Part B
create_airports = "
CREATE TABLE IF NOT EXISTS airports (
  aid INTEGER PRIMARY KEY,
  airportName TEXT,
  airportState TEXT,
  airportCode VARCHAR(255) DEFAULT 'ZZZ'
);
"
# Part D
create_lookup = "
CREATE TABLE IF NOT EXISTS conditions (
  cid INTEGER PRIMARY KEY,
  sky_condition TEXT,
  explanation VARCHAR(255) DEFAULT ' '
);
"
# Part E, F
create_wl_strikes = "
CREATE TABLE IF NOT EXISTS `wildlife.strikes` (
  iid INTEGER PRIMARY KEY,
  fid INTEGER,
  wlsize INTEGER,
  impact TEXT,
  altitude INTEGER,
  conditions INTEGER,
  CHECK (altitude >= 0),
  FOREIGN KEY (fid) REFERENCES flights (fid),
  FOREIGN KEY (conditions) REFERENCES conditions (cid),
  FOREIGN KEY (wlsize) REFERENCES wlsizes (wid)
);
"
# Part E
create_wlsizes = "
CREATE TABLE IF NOT EXISTS wlsizes (
  wid INTEGER PRIMARY KEY,
  size TEXT
);
"

# Use a function to loop through all sql statements to create the 5 needed tables
tables = c(create_flights, create_airports, create_lookup,
           create_wl_strikes, create_wlsizes)
create_tables = function(table_list){
  for (table in table_list){
    dbExecute(conn, table)
  }
}
create_tables(tables)

# Make sure all 5 tables are created and exist in the database
dbListTables(conn)
```
## Question 4. Part G: Check Table Constraints
```{r test_tables, echo=T, eval=F, warning=F}
# Check the needed columns exist for all tables
for (table in dbListTables(conn)){
  print (paste0("Table: ", table))
  if (table == "wildlife.strikes"){
    table = "`wildlife.strikes`"
  }
  columns = dbListFields(conn, table)
  print (paste0("Columns: ", columns))
  print ("-----------------------------------------")
}

# Check the data type of columns in flights table is correct
check_dtype = "
SHOW CREATE TABLE flights;
"
dbGetQuery(conn, check_dtype)

# Check the foreign key constraint in between the lookup table and wildlife.strikes table
check_fk = "
SELECT * FROM information_schema.table_constraints 
WHERE table_name = 'wildlife.strikes'
AND constraint_type = 'FOREIGN KEY'
"
dbGetQuery(conn, check_fk)

# Check the foreign key constraint in between the airports table and flights table
check_fk = "
SELECT * FROM information_schema.table_constraints 
WHERE table_name = 'flights'
AND constraint_type = 'FOREIGN KEY'
"
dbGetQuery(conn, check_fk)
```
## Question 5. Load the aircraft csv File
```{r load_csv, echo=F, warning=F}
# Load the aircraft csv file intoa data frame called bds.raw
csv_file = "BirdStrikesData-V4-SuF24.csv"
bds.raw = read.csv(file=csv_file)
head(bds.raw)
```
## Question 6. Populate the flights Table with Data From bds.raw
```{r populate_flights, echo=T, warning=F}
library(zoo)
# Create a function to extract a subset of dataset from bds.raw for the flights table
create_flights_df = function(df){
  flights_cols = c("flight_date", "airline", "aircraft",                               "heavy_flag")
  flights_df = df[, flights_cols]
  flights_df$fid = (1:nrow(flights_df))
  flights_df$originAirport = (1:nrow(flights_df))
  col_order = c("fid", "flight_date", "originAirport", "airline",                   "aircraft", "heavy_flag")
  flights_df = flights_df[, col_order]
  colnames(flights_df)[colnames(flights_df) == "flight_date"] =       "date"
  colnames(flights_df)[colnames(flights_df) == "airline"] =           "airlineName"
  colnames(flights_df)[colnames(flights_df) == "aircraft"] =          "aircraftType"
  colnames(flights_df)[colnames(flights_df) == "heavy_flag"] =        "isHeavy"
  return (flights_df)
}
flights_df = create_flights_df(bds.raw)

# Clean the data to fill missing dates, replace empty values in other columns as "unknown", and add 1 & 0 to isHeavy column to express T & F since boolean isn't supported by MySQL
flights_df$date = as.Date(strptime(flights_df$date, format="%m/%d/%y %H:%M"), format="%Y-%m-%d")
flights_df$date = zoo::na.locf(flights_df$date, na.rm = FALSE)
flights_df$airlineName[flights_df$airlineName == "" | flights_df$airlineName == " "] = "unknown"
flights_df$aircraftType[flights_df$aircraftType == "" | flights_df$aircraftType == " "] = "unknown"
flights_df$isHeavy[flights_df$isHeavy == "" | flights_df$isHeavy == " "] = "unknown"
flights_df$isHeavy = ifelse(flights_df$isHeavy == "Yes" | flights_df$isHeavy == "unknown", 1, 0)

# Create a function to batch insert the df data to the flights table
batch_insert = function(df, conn, query, batch_size = 1000) {
  row_num = nrow(df)
  for (idx in seq(1, row_num, by = batch_size)) {
    end = min(idx + batch_size - 1, row_num)
    batch = df[idx:end, ]
    values = apply(batch, 1, function(row) {
      formatted_values = sapply(row, function(value) {
        if (is.numeric(value)) {
          as.character(value)
        } else {
          paste0("'", gsub("'", "''", value), "'")
        }
      })
      paste0("(", paste(formatted_values, collapse = ", "), ")")
    })
    value_string = paste(values, collapse = ", ")
    batch_query = paste0(query, value_string)
    dbSendQuery(conn, batch_query)
  }
}

# Execute the function to fill data into the flights table
flights_query = "INSERT INTO flights (fid, date, originAirport, airlineName, aircraftType, isHeavy) VALUES "
batch_insert(flights_df, conn, flights_query)
```
## Question 6. Populate the airports table with data from bds.raw
```{r populate_airports, echo=T, warning=F}
# Create a function to extract a subset of dataset from bds.raw for the airports table
create_airports_df = function(df){
  airports_cols = c("dep_airport", "origin_state")
  airports_df = df[, airports_cols]
  airports_df$aid = (1:nrow(airports_df))
  col_order = c("aid", "dep_airport", "origin_state")
  airports_df = airports_df[, col_order]
  colnames(airports_df)[colnames(airports_df) == "dep_airport"] =     "airportName"
  colnames(airports_df)[colnames(airports_df) == "origin_state"] =    "airportState"
  return (airports_df)
}
airports_df = create_airports_df(bds.raw)

# Clean the data to replace missing airport states and empty values in airport name as "unknown"
airports_df$airportName[airports_df$airportName == "" | airports_df$airportName == " "] = "unknown"
airports_df$airportState[airports_df$airportState == "N/A"] = "unknown"

# Load the data frame into airports table
airports_query = "INSERT INTO airports (aid, airportName, airportState) VALUES "
batch_insert(airports_df, conn, airports_query)
```
## Question 6. Populate the conditions Table with Data From bds.raw
```{r populate_conditions, echo=T, warning=F}
# Insert data into the conditions lookup table
insert_conditions = "
INSERT INTO conditions VALUES
  (1, 'No Cloud', ' '),
  (2, 'Some Cloud', ' '),
  (3, 'Overcast', ' ');
"
dbExecute(conn, insert_conditions)
```
## Question 6. Populate the wlsizes Table with Data From bds.raw
```{r populate_wlsizes, echo=T, warning=F}
# Insert data into the wlsize lookup table
insert_wlsizes = "
INSERT INTO wlsizes VALUES
  (1, 'Small'),
  (2, 'Medium'),
  (3, 'Large');
"
dbExecute(conn, insert_wlsizes)
```
## Question 6. Populate the wildlife.strikes Table with Data From bds.raw
```{r populate_wildlife.strikes, echo=T, warning=F}
library(dplyr)
# Create a function to extract a subset of dataset from bds.raw for the wildlife.strikes table
create_wlstrikes_df = function(df){
  wlstrikes_cols = c("iid", "wildlife_size", "impact",                                   "altitude_ft", "sky_conditions")
  wlstrikes_df = df[, wlstrikes_cols]
  wlstrikes_df$fid = (1:nrow(wlstrikes_df))
  col_order = c("iid", "fid", "wildlife_size", "impact",                            "altitude_ft", "sky_conditions")
  wlstrikes_df = wlstrikes_df[, col_order]
  colnames(wlstrikes_df)[colnames(wlstrikes_df) ==                    "wildlife_size"] = "wlsize"
  colnames(wlstrikes_df)[colnames(wlstrikes_df) ==                    "altitude_ft"] = "altitude"
  colnames(wlstrikes_df)[colnames(wlstrikes_df) ==                    "sky_conditions"] = "conditions"
  return (wlstrikes_df)
}
wlstrikes_df = create_wlstrikes_df(bds.raw)

# Replacing empty values in wlsize, impact, and altitude column with the most common value in them
wlstrikes_df$wlsize[wlstrikes_df$wlsize == ""] = "Small"
wlstrikes_df$impact[wlstrikes_df$impact == ""] = "None"
wlstrikes_df$altitude[wlstrikes_df$altitude == ""] = 0

# Remove comma sign from the altitude column
wlstrikes_df$altitude = gsub(",", "", wlstrikes_df$altitude)

# Update the wlsize and conditions column to match the primary key value in the look up tables, which are wlsizes and conditions
wlstrikes_df = wlstrikes_df %>%
  mutate(wlsize = recode(wlsize, "Small" = 1, 
                                "Medium" = 2, "Large" = 3))
wlstrikes_df = wlstrikes_df %>%
  mutate(conditions = recode(conditions, "No Cloud" = 1, 
                             "Some Cloud" = 2, "Overcast" = 3))

# Load the data frame into airports table
wlstrikes_query = "INSERT INTO `wildlife.strikes` (iid, fid, wlsize, impact, altitude, conditions) VALUES "
batch_insert(wlstrikes_df, conn, wlstrikes_query)
```
## Question 7. Display the Table Contents
```{r display_tables, echo=F, eval=F, warning=F}
# Show the contents in all tables in the databse, in the form of data frames
all_tables = dbListTables(conn)
for (table in all_tables){
  if (table == "wildlife.strikes"){
    table = "`wildlife.strikes`"
  }
  query = paste0("SELECT * FROM ", table)
  df = dbGetQuery(conn, query)
  print (head(df))
}
```
### Enable the Foreign Key Constraint Again
```{r enable_fk_constraint, echo=T, warning=F}
# Now the tables are all properly created, enable the foreign key constraint again to support join statements in future sql queries
enable_fk = "
SET FOREIGN_KEY_CHECKS = 1;
"
dbExecute(conn, enable_fk)
```
## Question 8. Top States with Strikes
```{r states_with_most_strikes, echo=F, warning=F}
# Create a query statement to find the top 5 states with the most wildlife strikes incidents
library(kableExtra)
top_strikes = "
SELECT airportState, COUNT(aid) AS number_of_incidents
FROM airports
GROUP BY airportState
ORDER BY number_of_incidents DESC
LIMIT 5;
"

# Execute the query to get the result and organize it into a tabular format using kable() function
top_strikes_df = dbGetQuery(conn, top_strikes)
kable(top_strikes_df)
```
## Question 9. Analysis by Airport
```{r airport_analysis, echo=F, warning=F}
# Find the value of average number of wildlife strikes across all airports in the dataset
find_avg_incidents = "
SELECT AVG(incident_count) AS Avg_incident_count
FROM (
  SELECT COUNT(aid) AS incident_count
  FROM airports
  GROUP BY airportName
) AS incident_counts;
"
avg_incident_df = dbGetQuery(conn, find_avg_incidents)
avg_incident = avg_incident_df[1, 1]

# Create a query statement to find the first 5 airports with a below average number of wildlife strikes incident
below_avg_airports = paste0("
SELECT airportName, COUNT(aid) AS incident_count
FROM airports
GROUP BY airportName
HAVING incident_count < ", avg_incident, " LIMIT 5", ";")

# Execute the query to get the result and put it into tabular form using kable() function
below_avg_airport_df = dbGetQuery(conn, below_avg_airports)
kable(below_avg_airport_df)
```
## Question 10. Analysis by Year
```{r year_analysis, echo=F, warning=F}
# Add a new column called year in the flights table
add_year_column = "
ALTER TABLE flights
ADD COLUMN year INT DEFAULT 2000;
"
dbExecute(conn, add_year_column)

# Update the value in the year column based on the date column, in the flights table
update_year_column = "
UPDATE flights
SET year = (CASE
                WHEN date LIKE '2000%' THEN 2000
                WHEN date LIKE '2001%' THEN 2001
                WHEN date LIKE '2002%' THEN 2002
                WHEN date LIKE '2003%' THEN 2003
                WHEN date LIKE '2004%' THEN 2004
                WHEN date LIKE '2005%' THEN 2005
                WHEN date LIKE '2006%' THEN 2006
                WHEN date LIKE '2007%' THEN 2007
                WHEN date LIKE '2008%' THEN 2008
                WHEN date LIKE '2009%' THEN 2009
                WHEN date LIKE '2010%' THEN 2010
                WHEN date LIKE '2011%' THEN 2011
            END);
"
dbExecute(conn, update_year_column)

# Create a query statement to find the number of wildlife strikes per year for sizes of small, medium, and large wildlifes
find_wlsize_strikes = "
SELECT f.year,
       COUNT(CASE
                 WHEN s.size = 'Small' THEN w.iid
             END) AS small_wl_strikes,
       COUNT(CASE
                 WHEN s.size = 'Medium' THEN w.iid
             END) AS medium_wl_strikes,
       COUNT(CASE
                 WHEN s.size = 'Large' THEN w.iid
             END) AS large_wl_strikes
FROM `wildlife.strikes` AS w
INNER JOIN wlsizes AS s
ON (w.wlsize = s.wid)
INNER JOIN flights AS f
ON (w.fid = f.fid)
GROUP BY f.year
ORDER BY f.year;
"

# Execute the query to get the result and put it into tabular form using kable() function
wlsize_per_year = dbGetQuery(conn, find_wlsize_strikes)
kable(wlsize_per_year)
```
## Question 11. Trend by Year
```{r plot_year_trend, echo=T, warning=F}
# Write a for loop to get a list of years and list of total strikes from all sizes of wildlife for each year
all_total_strikes = c()
all_years = c()
for (row_idx in (1:nrow(wlsize_per_year))){
  year = wlsize_per_year[row_idx, 1]
  all_years = append(all_years, year)
  total_strikes = sum(wlsize_per_year[row_idx, ][2:4])
  all_total_strikes = append(all_total_strikes, total_strikes)
}

# Use the plot() function to create a scatter plot between years from 2000 to 2011 and number of wildlife strike incidents
plot(all_years, all_total_strikes, main="Total Wildlife Strikes per Year", xlab="Year", ylab="Wildlife Strikes", pch=19)
```
## Question 12. Create Stored Procedure
```{r create_stored_procedure, echo=T, warning=F}
# Create the table to log updates to strikes data
create_log_table = "
CREATE TABLE IF NOT EXISTS update_log(
  mod_id INTEGER AUTO_INCREMENT,
  modification TEXT,
  table_updated TEXT,
  update_time TIMESTAMP,
  original_value TEXT,
  PRIMARY KEY(mod_id)
);
"
dbExecute(conn, create_log_table)

# Create the stored procedure to update strikes data
create_procedure = "
CREATE PROCEDURE update_airport_code()
BEGIN
     UPDATE airports
     SET airportCode = 'LGA'
     WHERE airportName = 'LAGUARDIA NY';
END
"
dbExecute(conn, create_procedure)

# Create a trigger that updates the update_log table when the modification inside the stored procedure is executed
create_trigger = "
CREATE TRIGGER check_airportcode_update
AFTER UPDATE ON airports
FOR EACH ROW
BEGIN
     IF NEW.airportCode != OLD.airportCode THEN
      INSERT INTO update_log (modification, table_updated,                                        update_time, original_value)
      VALUES ('Update', 'airports', CURRENT_TIMESTAMP,                            OLD.airportCode);
     END IF;
END
"
dbExecute(conn, create_trigger)
```
## Question 12. Test the stored procedure
```{r test_procedure, echo=T, warning=F}
# Execute the stored procedure to update strikes data
dbExecute(conn, "CALL update_airport_code()")

# Print out update_log table to make sure the modification on the strike event is being logged
dbGetQuery(conn, "SELECT * FROM update_log LIMIT 10;")

# Print the rows from flights table to make sure the airport codes are changed to 'LGA' if the airport name is ' LAGUARDIA NY'
select_airports = "
SELECT aid, airportName, airportCode
FROM airports
WHERE airportName = 'LAGUARDIA NY'
LIMIT 10;
"
dbGetQuery(conn, select_airports)
```
### Disconnect from the database
```{r, disconnect_from_df, echo=T, warning=F}
# Disconnect from the MySQL databse server
dbDisconnect(conn)
```